Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-07-24 Manuscript/03-Simulation.Rmd
Revise the simulation section

# Simulation Study
In this section, we compare the performance of the proposed models to 3 other models: COSSO, adaptive COSSO ,mgcv. COSSO is xxx method. Adaptive COSSO is another method, mgcv is another method. 
To demonstrate the correctness and feasibility of the proposed SS-GAM,
we follow the logistic regression simulation introduced in Bai (Work in
Progress). We first generate $n=500$ training data sample with number
of predictors $p=4, 10, 50, 200$ respectively, where the predictors $X \sim MVN_{n\times p}(0, I_{P})$. We simulate the outcome data from three distributions, gaussian, poisson, and binomial with the identity link, logit link $g(x) = \log(\frac{x}{1-x})$, and log link $g(x) = log(x)$ functions to mimic the common use of generalized model. The outcome is simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
$$
for gaussian and binomial outcome and shrinked effects

$$
\mathbb{E}(Y) = g^{-1}(\sin(2\pi x_1) - 2\cos(2\pi x_2 -0.5) + 0.4(x_3-0.5) - 2(x_4^2 -0.3))
$$
for stable poisson outcome. For the Gaussian outcome where dispersion are required, the dispersion parameter was set to 1. We follow the same data generating functions in Bai. In this data generating process, we have $x_1, x_2, x_3, x_4$ as the active covariates, while the rest
covariates are inactive $f_j(x_j) = 0$ for $j = 4, \dots, p$. Another set of independent sample of size $n_{test}=1000$,($x_{new} ,y_{new}$), are created following the same data generating process, serving as the testing data. For mgcv and proposed models where choice of spline function are customizable, we choose to cubic regression spline with 10 basis. In COSSO and adaptive COSSO where the choice of spline are pre-determined, we set to have 10 bases. The corresponding number of parameter that we estimate with the model is 37, 82, 451 and 1801. We also run the simulation where the number of bases are increased to 25 deliberately. To evaluate the performance of the models, mean squared error, mean absolute error for gaussian model, AUC and misclassifcation rate are calculated for binomial model, and deviance are calculated for Poisson model, averaging across 50 simulations. The results are tabulated in Table 1, 2, 3.

>>>>

# Simulation Study

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for non-linear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \text{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence used the default choices. Both mgcv and proposed models allow customized smoothing functions and we chose the cubic regression spline. We controlled the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. 5-fold cross-validation were used for COSSO models, SB-GAM and the proposed models for tuning parameter selection based on the default selection criteria. 20 default candidates were examined for SB-GAM and the proposed models which allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].

Other related methods for high-dimensional GAMs also exist, notably the methods of sparse additive models by Ravikumar et al. [@Ravikumar2009], stochastic search term selection for GAM [@Scheipl2012]. However, we exclude these methods from current simulation study because of demonstrated inferior predictive performance compared to mgcv and scalability issues with increasing number of predictors. [@Scheipl2013]

## Monte Carlo Simulation Study
We follow the data generating process described in Bai [@Bai2021]. We first generated $n=500$ training data points with $p=4, 10, 50, 200$ predictors respectively, where the predictors $X$ are simulated from a multivariate normal distribution $\text{MVN}_{n\times p}(0, I_{P})$. We then simulated the outcome $y$ from three distribution distributions, gaussian, binomial, and poisson with the identity link, logit link $g(x) = \log(\frac{x}{1-x})$, and log link $g(x) = log(x)$. The mean of each outcome were simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
$$
for gaussian and binomial outcomes and shrinked effects

$$
\mathbb{E}(Y) = g^{-1}(\sin(2\pi x_1) - 2\cos(2\pi x_2 -0.5) + 0.4(x_3-0.5) - 2(x_4^2 -0.3))
$$
for  poisson outcomes. Gaussian outcomes required specification of dispersion, where we set the dispersion parameter to be 1. In this data generating process, we have $x_1, x_2, x_3, x_4$ as the active covariates, while the rest covariates are inactive, i.e. $f_j(x_j) = 0$ for $j = 4, \dots, p$. Another set of independent sample of size $n_{test}=1000$,($x_{new} ,y_{new}$), are created following the same data generating process, serving as the testing data. The dimensionality of the smooth functions are controlled as k=10. Hence, the corresponding number of coefficient parameters to be estimated is 37, 82, 451 and 1801. 
<!-- We also run the simulation where the number of bases are increased to 25 deliberately. -->

To evaluate the performance of the models, mean squared error, mean absolute error for gaussian model, AUC and misclassifcation rate are calculated for binomial model, and deviance are calculated for Poisson model, averaging across 50 simulations. The results are tabulated in Table 1, 2, 3.

### 2021-07-24 Manuscript/SS_GAM.tex
Revise the simulation section

Alternative to the sampling algorithms for fitting Bayesian models,
optimization algorithms focus on the maximum a posteriori (MAP)
estimates and speed up the process at the cost of uncertainty measures.
The earliest work for fitting spike-and-slab models using optimization
algorithm is EMVS \citep{Rockova2014a}. \citet{Rockova2014a} proposed an
expectation-maximization (EM) based algorithm to fit models that use
continuous mixture normal priors. In the E step, the latent variables
\(\gamma\) are treated as the missing data, and calculate the posterior
mean of the density function; in the M step, a ridge estimator was used
to update the coefficients, and \(\sigma^2, \theta\) are updated
accordingly. The same authors \citep{Rockova2018b, Rockova2018} further
combined the EM algorithm with coordinate descent (CD) algorithm for
fitting SSL models. Yi and his group incorporated the SSL prior in the
EM-Iterative Weighted Least Square(IWLS) algorithm and proposed another
EM algorithm, EM-cyclic coordinate descent, based on the Lasso penalty
SSL exhibit. Such applications were used in generalized linear
models\citep{Tang2017a}, cox models \citep{Tang2017} and their grouped
counterparts\citep{Tang2018, Tang2019}. Both EM based algorithms provide
deterministic solutions, which becomes a more popular property as
reproducible results. Bai et al. \citep{Bai2020} provides an up-to-date
summary of SSL methods and theoretical discussion.

@@

In this section, we compare the performance of the proposed models to 3
other models: COSSO, adaptive COSSO ,mgcv. COSSO is xxx method. Adaptive
COSSO is another method, mgcv is another method. To demonstrate the
correctness and feasibility of the proposed SS-GAM, we follow the
logistic regression simulation introduced in Bai (Work in Progress). We
first generate \(n=500\) training data sample with number of predictors
\(p=4, 10, 50, 200\) respectively, where the predictors
\(X \sim MVN_{n\times p}(0, I_{P})\). We simulate the outcome data from
three distributions, gaussian, poisson, and binomial with the identity
link, logit link \(g(x) = \log(\frac{x}{1-x})\), and log link
\(g(x) = log(x)\) functions to mimic the common use of generalized
model. The outcome is simulated via the following function \[
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
\] for gaussian and binomial outcome and shrinked effects

>>>>

Alternative to the sampling algorithms for fitting Bayesian models,
optimization algorithms focus on the maximum a posteriori (MAP)
estimates and speed up the process at the cost of uncertainty measures.
The earliest work for fitting spike-and-slab models using optimization
algorithm is EMVS \citep{Rockova2014a}. Ro\{\v{c}\}kov\{'\{a\}\} and
George\citep{Rockova2014a} proposed an expectation-maximization (EM)
based algorithm to fit models that use continuous mixture normal priors.
In the E step, the latent variables \(\gamma\) are treated as the
missing data, and calculate the posterior mean of the density function;
in the M step, a ridge estimator was used to update the coefficients,
and \(\sigma^2, \theta\) are updated accordingly. The same authors
\citep{Rockova2018b, Rockova2018} further combined the EM algorithm with
coordinate descent (CD) algorithm for fitting SSL models. Yi and his
group incorporated the SSL prior in the EM-Iterative Weighted Least
Square(IWLS) algorithm and proposed another EM algorithm, EM-cyclic
coordinate descent, based on the Lasso penalty SSL exhibit. Such
applications were used in generalized linear models\citep{Tang2017a},
cox models \citep{Tang2017} and their grouped
counterparts\citep{Tang2018, Tang2019}. Both EM based algorithms provide
deterministic solutions, which becomes a more popular property as
reproducible results. Bai et al. \citep{Bai2020} provides an up-to-date
summary of SSL methods and theoretical discussion.

@@

In this section, we compare the performance of the proposed models to
four alternative models: component selection and smoothing operator
(COSSO) \citep{Zhang2006GAM}, adaptive COSSO \citep{Storlie2011},
generalized additive models with automatic smoothing \citep{Wood2011},
SB-GAM \citep{Bai2021}. COSSO is one of the earliest smoothing spline
models that consider sparsity-smoothness penalty. Adaptive COSSO
improved upon COSSO by using adaptive weight for penalties such that the
penalty of each functional component are different for extra
flexibility. Generalized additive models with automatic smoothing,
hereafter \textit{mgcv}, is one of the most popular models for
non-linear effect interpolation and prediction. SB-GAM is the first
spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R
package \texttt{cosso} \citep{cosso} 2.1-1, generalized additive models
with automatic smoothing with R package \texttt{mgcv} 1.8-31, SB-GAM
with R package \text{sparseGAM} 1.0. All computation was conducted on a
high-performance 64-bit Linux platform with 48 cores of 2.70GHz
eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R
3.6.2 \citep{R}.

Other related methods for high-dimensional GAMs also exist, notably the
methods of sparse additive models by Ravikumar et al.
\citep{Ravikumar2009}, stochastic search term selection for GAM
\citep{Scheipl2012}. However, we exclude these methods from current
simulation study because of demonstrated inferior predictive performance
compared to mgcv and scalability issues with increasing number of
predictors. \citep{Scheipl2013}

\hypertarget{monte-carlo-simulation-study}{%
\subsection{Monte Carlo Simulation
Study}\label{monte-carlo-simulation-study}}

xxx method. Adaptive COSSO is another method, mgcv is another method. To
demonstrate the correctness and feasibility of the proposed SS-GAM, we
follow the logistic regression simulation introduced in Bai (Work in
Progress). We first generate \(n=500\) training data sample with number
of predictors \(p=4, 10, 50, 200\) respectively, where the predictors
\(X \sim MVN_{n\times p}(0, I_{P})\). We simulate the outcome data from
three distributions, gaussian, poisson, and binomial with the identity
link, logit link \(g(x) = \log(\frac{x}{1-x})\), and log link
\(g(x) = log(x)\) functions to mimic the common use of generalized
model. The outcome is simulated via the following function \[
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
\] for gaussian and binomial outcome and shrinked effects
